---
title: "STA 4210 HW 3"
author: "Yansheng Luo"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE)
```

1.(R) In this problem, you will be working with the `cars` dataset, which contains two variables:
    - `speed`: The speed of the car (in mph)
    - `dist`: The stopping distance (in feet)

```{r}
library(car)
car.reg <- lm(dist~speed,data = cars)

e <- rstudent(car.reg)
shapiro.test(e) # Shapiro-Wilk
ncvTest(car.reg) # Breusch-Pagan
```


a. (5pt) Based on the results of the significance test, there might be violation to the Homogeneity assumption and Normal assumption. Use the `boxcox` function from the `MASS` package to find the optimal value of the Box-Cox transformation parameter ($\lambda$).




b. (5pt) Apply the transformation to the response variable using the optimal $\lambda$ and refit the model using the transformed data.


c. (10pt) Check ALL model assumptions by significance tests with $\alpha = 0.05$ after applying the transformation. Are there any potential violations? 

d. (5pt) For the refitted model with the transformed response, compute simultaneous confidence intervals for the mean transformed stopping distance at speed = 20 mph and 24 mph using the Bonferroni method with at least 95% coverage. You can use the `predict` function.




2.(R) In this problem, you will work with matrices in R. You can use the following R functions to complete the tasks:
    - `matrix()` to generate a matrix
    - `%*%` for matrix multiplication
    - `det()` to calculate the determinant
    - `qr()` to find the rank of a matrix
    - `t()` for the transpose of a matrix





a. (5pt) Let $A$ and $B$ be the following matrices:
$$
A = \begin{pmatrix}
2 & 1 & 0 \\
-1 & 3 & 4 \\
5 & 6 & -2
\end{pmatrix}
$$

$$
B = \begin{pmatrix}
1 & 2 \\
-1 & 0 \\
3 & 4
\end{pmatrix}
$$
Multiply matrix $A$ by matrix $B$ (i.e., compute $AB$).




b. (5pt) Calculate the transpose of matrix $A$ (denoted $A^T$). Is matrix $A^TA$ symmetric? Verify this by checking whether it is equal to its transpose.



c. (5pt) Consider the matrix $C$:
$$
C = \begin{pmatrix}
3 & 1 \\
5 & 7
\end{pmatrix}
$$Calculate the determinant of matrix $C$. What does the determinant tell you about the matrix?



d. (5pt) Using matrix $A$ from part (a). Calculate the rank of matrix $A$. Is matrix $A$ full-rank (i.e., does it have the maximum rank)? *R Hint*: You can use `qr()` function obtain the rank directly, or use `eigen()` function to find the eigenvalues and rank is the number of non-zero eigenvalues.



e. (5pt) If a square matrix is full-rank, its inverse can be computed. Consider the following matrix $D$:
$$
D = \begin{pmatrix}
4 & 7 \\
2 & 6
\end{pmatrix}
$$Check if matrix $D$ is full-rank. If $D$ is full-rank, calculate its inverse. *R Hint*: Use `solve()` to calculate the inverse of a matrix. Make sure to check the rank first.
