---
title: "STA 4210 HW 3"
author: "Yansheng Luo"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE)
```

1.(R) In this problem, you will be working with the `cars` dataset, which contains two variables:
    - `speed`: The speed of the car (in mph)
    - `dist`: The stopping distance (in feet)

```{r}
library(car)
car.reg <- lm(dist~speed,data = cars)

e <- rstudent(car.reg)
shapiro.test(e) # Shapiro-Wilk
ncvTest(car.reg) # Breusch-Pagan
```


a. (5pt) Based on the results of the significance test, there might be violation to the Homogeneity assumption and Normal assumption. Use the `boxcox` function from the `MASS` package to find the optimal value of the Box-Cox transformation parameter ($\lambda$).

## Solutions:
The Box--Cox transformation for a positive response variable $Y$ is defined as
$$
Y^{(\lambda)} =
\begin{cases}
\dfrac{Y^\lambda - 1}{\lambda}, & \lambda \neq 0 \\
\log(Y), & \lambda = 0
\end{cases}
$$

The optimal value of $\lambda$ is chosen to maximize the log-likelihood function.
```{r question 1 part a}
library(MASS)

# Box-Cox on the ORIGINAL model
bc <- boxcox(car.reg, lambda = seq(-2, 2, by = 0.01), plotit = TRUE)

# Optimal lambda = argmax of log-likelihood
lambda_opt <- bc$x[which.max(bc$y)]
lambda_opt
```

The Box--Cox analysis suggests an optimal transformation parameter of
$\hat{\lambda} = 0.43$. Since this value is close to $0.5$, the transformation
is approximately a square-root transformation, which may help stabilize the
variance and improve the normality of the residuals.

---

b. (5pt) Apply the transformation to the response variable using the optimal $\lambda$ and refit the model using the transformed data.

## Solution:
The transformed response variable is computed using the Box--Cox formula

$$
Y^{(\lambda)} =
\frac{Y^\lambda - 1}{\lambda}, \quad \lambda \neq 0.
$$

Substituting $\hat{\lambda} = 0.43$,

$$
Y^{(0.43)} =
\frac{Y^{0.43} - 1}{0.43}.
$$

```{r question 1 part b}
lambda_opt <- 0.43

# Transform response
dist_bc <- (cars$dist^lambda_opt - 1) / lambda_opt

# Refit regression model
car.reg.bc <- lm(dist_bc ~ speed, data = cars)

summary(car.reg.bc)
```
From the output:
The intercept = 1.05004
slope = 0.50538
Thus:
$$
\widehat{Y^{(0.43)}} =
1.05004 + 0.50538({x}).
$$
The response variable was transformed using the optimal Box--Cox parameter
$\hat{\lambda}=0.43$, and the regression model was refit using the transformed
response. The fitted model shows a significant positive relationship between
speed and the transformed stopping distance ($p < 0.001$).

---

c. (10pt) Check ALL model assumptions by significance tests with $\alpha = 0.05$ after applying the transformation. Are there any potential violations? 

## Solution:
After applying the Box–Cox transformation with 
$\hat{\lambda} = 0.43$, the regression assumptions were evaluated using significance tests at $\alpha=0.05$.

The following tests were performed based on the previous HW:

1.Shapiro–Wilk test (normality)

2.Runs test (randomness)

3.Durbin–Watson test (independence)

4.Levene’s test (homogeneity of variance)

5.Breusch–Pagan / Cook–Weisberg test (heteroscedasticity)

6.Lack-of-fit test (model adequacy / linearity)

```{r question 1 part c}
if(!require(randtests)) install.packages("randtests")
if(!require(lmtest)) install.packages("lmtest")
if(!require(car)) install.packages("car")

library(randtests)
library(lmtest)
library(car)


# Studentized residuals
e_bc <- rstudent(car.reg.bc)

### 1. Shapiro-Wilk test (Normality)
shapiro.test(e_bc)

### 2. Runs test (Independence/if the residual is random)
runs.test(e_bc)

### 3. Durbin-Watson test (Independence)
durbinWatsonTest(car.reg.bc)

### 4. Levene's Test (Homogeneity)
leveneTest(e_bc ~ as.factor(cars$speed))

### 5. Breusch-Pagan / Cook-Weisberg (Homogenetiy)
bptest(car.reg.bc)

### 6. Lack-of-Fit Test (linearity)
# Compare linear vs saturated model
car.reg.full <- lm(dist_bc ~ factor(speed), data = cars)
anova(car.reg.bc, car.reg.full)
```

\textbf{Shapiro--Wilk test (Normality)}

$$
H_0: \text{Residuals are normally distributed}
$$

The p-value is $0.3403 > 0.05$, so we fail to reject $H_0$. There is no evidence
against the normality assumption.

\textbf{Runs test (Independence/Randomness)}

$$
H_0: \text{Residuals occur in random order}
$$

The p-value is $0.5676 > 0.05$, so we fail to reject $H_0$. The residuals appear
to occur randomly and are independent.

\textbf{Durbin--Watson test (Independence)}

$$
H_0: \text{Errors are independent}
$$

The p-value is $0.796 > 0.05$, so we fail to reject $H_0$. There is no evidence
of autocorrelation among the residuals.

\textbf{Levene's test (Homogeneity of variance)}

$$
H_0: \sigma^2_1 = \sigma^2_2 = \cdots = \sigma^2_k
$$

The p-value is $0.2849 > 0.05$, so we fail to reject $H_0$. The variances appear
to be homogeneous.

\textbf{Breusch--Pagan test (Constant variance)}

$$
H_0: \sigma^2 \text{ is constant}
$$

The p-value is $0.7382 > 0.05$, so we fail to reject $H_0$. There is no evidence
of heteroscedasticity.

\textbf{Lack-of-fit test (Model adequacy)}

$$
H_0: \text{The linear model is adequate (no lack of fit)}
$$

The p-value is $0.5177 > 0.05$, so we fail to reject $H_0$. The linear model
appears to provide an adequate fit to the data.

\textbf{Overall Conclusion}

Since all p-values are greater than $\alpha = 0.05$, there is no evidence of
violations of the regression assumptions after applying the Box--Cox
transformation. Therefore, the transformed model satisfies the assumptions of
normality, independence, homogeneity of variance, randomness, and linear model
adequacy.

---

d. (5pt) For the refitted model with the transformed response, compute simultaneous confidence intervals for the mean transformed stopping distance at speed = 20 mph and 24 mph using the Bonferroni method with at least 95% coverage. You can use the `predict` function.

## Solution:

```{r question 1 part d}
# Bonferroni simultaneous confidence intervals for the MEAN transformed response
# at speed = 20 and 24 mph (coverage >= 95%)

newdata <- data.frame(speed = c(20, 24))
g <- nrow(newdata)          # number of intervals
alpha <- 0.05
level_bonf <- 1 - alpha/g   # Bonferroni per-interval confidence level = 0.975

pred_bonf <- predict(car.reg.bc,
                     newdata = newdata,
                     interval = "confidence",
                     level = level_bonf)

# Display results with speeds
cbind(newdata, pred_bonf)
```
We construct simultaneous confidence intervals for the mean transformed
stopping distance at speeds 20 mph and 24 mph using the Bonferroni method
with overall confidence level at least 95\%.

Since two intervals are required ($g=2$) and $\alpha = 0.05$, the Bonferroni
per-interval confidence level is

$$
1 - \frac{\alpha}{g}
=
1 - \frac{0.05}{2}
=
0.975.
$$

Thus, each individual confidence interval is constructed at the 97.5\%
confidence level to guarantee at least 95\% simultaneous coverage.

Using the fitted transformed model and the \texttt{predict} function, the
simultaneous confidence intervals for the mean transformed stopping distance
are:

For speed $=20$ mph:

$$
(10.4104,\; 11.9048)
$$

For speed $=24$ mph:

$$
(12.0996,\; 14.2586)
$$

These intervals provide at least 95\% simultaneous confidence that the true
mean transformed stopping distances at speeds 20 mph and 24 mph lie within
their respective intervals.

---

2.(R) In this problem, you will work with matrices in R. You can use the following R functions to complete the tasks:
    - `matrix()` to generate a matrix
    - `%*%` for matrix multiplication
    - `det()` to calculate the determinant
    - `qr()` to find the rank of a matrix
    - `t()` for the transpose of a matrix





a. (5pt) Let $A$ and $B$ be the following matrices:
$$
A = \begin{pmatrix}
2 & 1 & 0 \\
-1 & 3 & 4 \\
5 & 6 & -2
\end{pmatrix}
$$

$$
B = \begin{pmatrix}
1 & 2 \\
-1 & 0 \\
3 & 4
\end{pmatrix}
$$
Multiply matrix $A$ by matrix $B$ (i.e., compute $AB$).

## Solution:

```{r question 2 part a}
# Define matrices A and B

A <- matrix(c(2, -1, 5,
              1,  3, 6,
              0,  4, -2),
            nrow = 3, byrow = FALSE)

B <- matrix(c(1, -1, 3,
              2,  0, 4),
            nrow = 3, byrow = FALSE)

# Matrix multiplication AB
AB <- A %*% B

AB
```

We are given the matrices

$$
A =
\begin{pmatrix}
2 & 1 & 0 \\
-1 & 3 & 4 \\
5 & 6 & -2
\end{pmatrix},
\qquad
B =
\begin{pmatrix}
1 & 2 \\
-1 & 0 \\
3 & 4
\end{pmatrix}.
$$

The matrix product $AB$ is computed by multiplying rows of $A$
with columns of $B$.

$$
AB =
\begin{pmatrix}
2 & 1 & 0 \\
-1 & 3 & 4 \\
5 & 6 & -2
\end{pmatrix}
\begin{pmatrix}
1 & 2 \\
-1 & 0 \\
3 & 4
\end{pmatrix}
=
\begin{pmatrix}
1 & 4 \\
8 & 14 \\
-7 & 2
\end{pmatrix}.
$$

Thus, the matrix product is

$$
AB =
\begin{pmatrix}
1 & 4 \\
8 & 14 \\
-7 & 2
\end{pmatrix}.
$$

---

b. (5pt) Calculate the transpose of matrix $A$ (denoted $A^T$). Is matrix $A^TA$ symmetric? Verify this by checking whether it is equal to its transpose.
## solution:
```{r question 2 part b}
# Matrix A from part (a)
A <- matrix(c(2, -1, 5,
              1,  3, 6,
              0,  4, -2),
            nrow = 3, byrow = FALSE)

# Transpose of A
A_T <- t(A)
A_T

# Compute A^T A
ATA <- A_T %*% A
ATA

# Check if symmetric: compare with its transpose
is_symmetric <- all(ATA == t(ATA))
is_symmetric
```
The transpose of matrix $A$ is obtained by interchanging rows and columns.

Given

$$
A =
\begin{pmatrix}
2 & 1 & 0 \\
-1 & 3 & 4 \\
5 & 6 & -2
\end{pmatrix},
$$

the transpose is

$$
A^T =
\begin{pmatrix}
2 & -1 & 5 \\
1 & 3 & 6 \\
0 & 4 & -2
\end{pmatrix}.
$$

Next, we compute $A^TA$:

$$
A^TA =
\begin{pmatrix}
30 & 29 & -14 \\
29 & 46 & 0 \\
-14 & 0 & 20
\end{pmatrix}.
$$

A matrix is symmetric if it equals its transpose. Since

$$
(A^TA)^T = A^TA,
$$

the matrix $A^TA$ is symmetric. This was also verified in R by checking
\texttt{all(ATA == t(ATA))}, which returned TRUE.

---


c. (5pt) Consider the matrix $C$:
$$
C = \begin{pmatrix}
3 & 1 \\
5 & 7
\end{pmatrix}
$$Calculate the determinant of matrix $C$. What does the determinant tell you about the matrix?

```{r question 2 part c}
# Define matrix C
C <- matrix(c(3, 5,
              1, 7),
            nrow = 2, byrow = FALSE)

# Determinant of C
det_C <- det(C)
det_C
```


We are given the matrix

$$
C =
\begin{pmatrix}
3 & 1 \\
5 & 7
\end{pmatrix}.
$$

The determinant of a $2 \times 2$ matrix

$$
\begin{pmatrix}
a & b \\
c & d
\end{pmatrix}
$$

is computed as

$$
\det(C) = ad - bc.
$$

Thus,

$$
\det(C) = (3)(7) - (1)(5) = 21 - 5 = 16.
$$

Since the determinant is nonzero ($\det(C)=16 \neq 0$), the matrix is
invertible and has full rank. This means its columns (and rows) are
linearly independent, and the matrix has an inverse.

---

d. (5pt) Using matrix $A$ from part (a). Calculate the rank of matrix $A$. Is matrix $A$ full-rank (i.e., does it have the maximum rank)? *R Hint*: You can use `qr()` function obtain the rank directly, or use `eigen()` function to find the eigenvalues and rank is the number of non-zero eigenvalues.

```{r question 2 part d}
# Matrix A from part (a)
A <- matrix(c(2, -1, 5,
              1,  3, 6,
              0,  4, -2),
            nrow = 3, byrow = FALSE)

# Rank using QR decomposition
rank_A <- qr(A)$rank
rank_A

# Check full-rank (maximum rank for 3x3 is 3)
is_full_rank <- (rank_A == min(dim(A)))
is_full_rank
```

Using matrix
$$
A =
\begin{pmatrix}
2 & 1 & 0 \\
-1 & 3 & 4 \\
5 & 6 & -2
\end{pmatrix},
$$
we compute the rank of $A$ in R using the QR decomposition via
\texttt{qr(A)\$rank}.

The result is
$$
\operatorname{rank}(A) = 3.
$$

Since $A$ is a $3\times 3$ matrix, its maximum possible rank is
$$
\min(3,3)=3.
$$
Because $\operatorname{rank}(A)=3$, matrix $A$ is full-rank. Equivalently,
its columns are linearly independent, and $A$ is invertible.

---

e. (5pt) If a square matrix is full-rank, its inverse can be computed. Consider the following matrix $D$:
$$
D = \begin{pmatrix}
4 & 7 \\
2 & 6
\end{pmatrix}
$$Check if matrix $D$ is full-rank. If $D$ is full-rank, calculate its inverse. *R Hint*: Use `solve()` to calculate the inverse of a matrix. Make sure to check the rank first.

```{r question 2 part e}
# Define matrix D
D <- matrix(c(4, 2,
              7, 6),
            nrow = 2, byrow = FALSE)

# Rank of D
rank_D <- qr(D)$rank
rank_D

# Check if full-rank
is_full_rank <- (rank_D == min(dim(D)))
is_full_rank

# Inverse of D (only if full-rank)
D_inv <- solve(D)
D_inv
```

We are given

$$
D =
\begin{pmatrix}
4 & 7 \\
2 & 6
\end{pmatrix}.
$$

First, we check whether $D$ is full-rank. The rank of a $2\times2$ matrix
can be at most 2. Using R, we obtain

$$
\operatorname{rank}(D)=2.
$$

Since the rank equals the maximum possible rank, matrix $D$ is full-rank
and therefore invertible.

The inverse of a $2\times2$ matrix

$$
\begin{pmatrix}
a & b \\
c & d
\end{pmatrix}
$$

is given by

$$
\frac{1}{ad-bc}
\begin{pmatrix}
d & -b \\
-c & a
\end{pmatrix}.
$$

The determinant of $D$ is

$$
\det(D) = (4)(6) - (7)(2) = 24 - 14 = 10.
$$

Thus,

$$
D^{-1}
=
\frac{1}{10}
\begin{pmatrix}
6 & -7 \\
-2 & 4
\end{pmatrix}
=
\begin{pmatrix}
0.6 & -0.7 \\
-0.2 & 0.4
\end{pmatrix}.
$$

Therefore, matrix $D$ is full-rank and its inverse exists.

